---
title: "Chapter 5"
output: html_document
date: '2023-06-03'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE}
#install.packages("Lahman")
library(dplyr)
library(nycflights13)
library(ggplot2)
```

### 5.5.2 Exercises
> 1. Currently dep_time and sched_dep_time are convenient to look at, but hard to compute with because they’re not really continuous numbers. Convert them to a more convenient representation of number of minutes since midnight.

[Using 'flights' dataset, select only dep_time and create one column 'dep_time_minutes' which has the number of minutes since midnight. Also note that the original dataset has midnight stored as 1440 which needs to be changed to 0.]

```{r}
flights %>% 
select(dep_time) %>% 
mutate(
  dep_time_minutes = (dep_time %/% 100 * 60 + dep_time %% 100) %% 1440
)
```



> 2. Compare air_time with arr_time - dep_time. What do you expect to see? What do you see? What do you need to do to fix it?

[Using 'flights' dataset, create new columns to convert arr_time and dep_time to minutes after midnight as arr_time_minutes and dep_time_minutes and calc_air_time as the difference. select only the new columns and air_time to compare]

```{r}
flights %>%
  mutate(
    arr_time_minutes = (arr_time %/% 100 * 60 + arr_time %% 100) %% 1440,
    dep_time_minutes = (dep_time %/% 100 * 60 + dep_time %% 100) %% 1440,
    calc_air_time = arr_time_minutes - dep_time_minutes
  ) %>%
select(arr_time_minutes, dep_time_minutes, calc_air_time, air_time)
```


> 3. Compare dep_time, sched_dep_time, and dep_delay. How would you expect those three numbers to be related?

[Use 'flights' dataset. Create new variables 'dep_time_minutes' and 'sched_dep_time_minutes' which represent minutes since midnight, 'calc_dep_delay' and a 'diff' showing difference of 'calc_dep_delay' and 'dep_delay'. Select only the pertinent columns, then sort with largest differences at the top.]
```{r}
flights %>%
  mutate(
    dep_time_minutes = (dep_time %/% 100 * 60 + dep_time %% 100) %% 1440,
    sched_dep_time_minutes = (sched_dep_time %/% 100 * 60 + sched_dep_time %% 100) %% 1440,
    calc_dep_delay = dep_time_minutes - sched_dep_time_minutes,
    diff = calc_dep_delay - dep_delay
  ) %>% select(dep_time_minutes, sched_dep_time_minutes, calc_dep_delay, diff) %>% arrange(desc(diff))
```


> 4. Find the 10 most delayed flights using 3 ranking functions [min_rank, row_number, dense_rank].

[Use 'flights' dataset; select the year, month, day, flight, tailnum, origin, dest, and dep_delay create 3 'dep_delay' rank columns, show only those less than 10, sorted by min_rank]
```{r}
flights %>%
  select(year, month, day, flight, tailnum, origin, dest, dep_delay) %>%
  mutate(dep_delay_min_rank = min_rank(desc(dep_delay)),
         dep_delay_row_number = row_number(desc(dep_delay)),
         dep_delay_dense_rank = dense_rank(desc(dep_delay))) %>%
  filter(!(dep_delay_min_rank > 10 | dep_delay_row_number > 10 | dep_delay_dense_rank > 10)) %>%
  arrange(dep_delay_min_rank)
```

>4. How are the functions min_rank() and dense_rank() different?

They both assigned the same value to "ties"; however they differ in how they handle the subsequent row.
min_rank() will have gaps as it assigns the next (non-tied) row the minimum integer value place greater than the number of values less that it (ex. 1, 2, 2, 2, 5)
dense_rank() will not have gaps as it assigns the next (non-tied) row the subsequent integer after the the tied ranking (ex. 1, 2, 2, 2, 3)

> 5. What does 1:3 + 1:10 return? Why?

It returns a warning "...longer object length is not a multiple of shorter object length". When vectors are combined in R if they are different length r "recycles" the smaller to match the length of the longer. However, if the multiple of the smaller does not match that of the longer a warning message is displayed.

> 6. What are the base trigonometric functions that R provide? 

* cos(x) or cospi(x)
* sin(x) or sinpi(x)
* tan(x) or tanpi(x)

> 6b. Using cosine of \(\frac{\pi}{6}\) show the two types are the same.

```{r}
identical(cos(pi/6), cospi(1/6))
```





## 5.6 Grouped Summaries with summarise()


What is the mean departure delay? 

[Use 'flights' dataset; use the summarise function and remove NAs]
```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
```

What is the mean departure delay by day? 

[Use 'flights' dataset; Use summarise function, remove NAs]
```{r}
group_by(flights, year, month, day) %>%
  summarise(delay = mean(dep_delay, na.rm = TRUE))
```

Create a new data frame called 'delay_v1' with the following analysis:

1. Group flights by destination
2. Summarise to compute average distance, average delay, and number of flights.
3. Filter to remove noise (include only the destinations with more than 20 flights) and remove Honolulu airport, which is almost twice as far away as the next closest airport.

Use 'flights' dataset:

```{r}
delay_v1 <- flights %>%
  group_by(dest) %>%
  summarise(dist = mean(distance, na.rm = TRUE),
            delay = mean(arr_delay, na.rm = TRUE),
            count = n(),) %>%
  filter(count > 20, dest != "HNL")
```


Plot delay vs. dist as a scatter with the size of the point based on count. Make it transparent (alpha = 1/3). Add a smoothed curve without the standard error.

Use the 'delay' dataset (previously created with the 'flights' dataset):

```{r, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)
```

```{r}
not_cancelled <- flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay))

not_cancelled %>%
  group_by(year, month, day) %>%
  summarise(mean = mean(dep_delay), .groups = "drop_last")
```

There is a subset of the 'flights' data set called 'not_cancelled'. 

[For context, it has filtered out any cancelled flights by using null values in 'dep_delay' and 'arr_delay' as proxies for cancelled flight]

Find out information about delays by tailnum by summarizing by tailnum and taking the mean of the arrival delay.

[This will produce the "1st" version of the "delays" dataset. Tailnum and average delay]

```{r}
delays_v1 <- not_cancelled %>%
  group_by(tailnum) %>%
  summarise(
    delay = mean(arr_delay)
  )

delays_v1
```

The 'delays_version1' dataset has a the average arrival delay by tailnum. Use the dataset to create a frequency polygon for the average delay with a binwidth of 10.
```{r}
ggplot(data = delays, mapping = aes(x = delay)) +
  geom_freqpoly(binwidth = 10)
```

The 'delay_version1' dataset shows us that some planes have an average delay of 300 minutes! Let's recreate the delays dataset [as 'delays_v2' from the 'not_cancelled' dataset, which has the same structure as the original 'flights' dataset. This time remove the nulls from 'arr_delay' and also create the count of flights.



```{r}
delays <- not_cancelled %>%
  group_by(tailnum) %>%
  summarise(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )
```

Now create a scatter plot of the average delay and flight count:  Delay vs. Count (n). [Use 'delays' dataset which is at the grain of tailnum in the 'delays' dataset. Given the number of points let's set alpha to 1/10]

```{r}
ggplot(data = delays, mapping = aes(x = n, y = delay)) +
  geom_point(alpha = 1/10)
```

Let's plot the information again, but let's only look at airplanes that had more than 25 flights. [Essentially removing the planes with few flights]
```{r}
delays %>%
  filter(n > 25) %>%
  ggplot(mapping = aes(x = n, y = delay)) +
  geom_point(alpha = 1/10)
```

```{r}
library(Lahman)
batting <- as_tibble(Lahman::Batting)
```

With the 'batting' dataset create a batting average column 'ba' (Hits 'H' / At Bats 'AB') and the At Bats 'ab' column removing "na"s all by PlayerID 
```{r}
batters <- batting %>%
  group_by(playerID) %>% 
  summarise(
    ba = sum(H, na.rm = TRUE) / sum(AB, na.rm = TRUE),
    ab = sum(AB, na.rm = TRUE)
    )
```


Using the 'batters' dataset plot a scatter plot of batting average 'ba' vs. at bats 'ab' for players with more than 100 at bats. Also, add a smoothed line showing the relationship (no need to add standard error)
```{r}
batters %>% 
  filter(ab > 100) %>% 
  ggplot(mapping = aes(x = ab, y = ba)) +
    geom_point() + 
    geom_smooth(se = FALSE)
```
GAM = Generalized Additive Model




Using the non_cancelled flights (just the subset of flights that hasn't been cancelled) get the average arrival delay by day as well as the average positive delay
```{r}
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(
    avg_delay1 = mean(arr_delay),
    avg_delay2 = mean(arr_delay[arr_delay > 0]),
    .groups = "drop_last"
  )
```

Using the 'not_cancelled' dataset find the standard deviation of distance by destination and sort from largest standard deviation to smallest.
```{r}
not_cancelled %>% 
  group_by(dest) %>% 
  summarise(distance_sd = sd(distance)) %>% 
  arrange(desc(distance_sd))
```

Using 'not_cancelled' dataset find the first and last departure time of each day. (using min and max functions in a group/summarise)
```{r}
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(
    first = min(dep_time),
    last = max(dep_time),
    .groups = "drop_last"
  )
```

Using 'not_cancelled' dataset find the first and last departure time of each day (using first and last functions in a group/summarise)
```{r}
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(
    first_dep = first(dep_time), 
    last_dep = last(dep_time),
    .groups = "drop_last"
  )
```

Using 'not_cancelled' dataset find the first and last departure time of each day. (using group_by, mutate, filter)
```{r}
not_cancelled %>% 
  group_by(year, month, day) %>% 
  mutate(r = min_rank(desc(dep_time))) %>% 
  filter(r %in% range(r))
```
Which destinations have the most carriers?
```{r}
not_cancelled %>% 
  group_by(dest) %>% 
  summarise(carriers = n_distinct(carrier)) %>% 
  arrange(desc(carriers))
```
Which destinations have the most flights? (filter and sort; use the n() function)
```{r}
not_cancelled %>% 
  group_by(dest) %>% 
  summarise(n = n()) %>%
  arrange(desc(n))
```

Which destinations have the most flights? (filter and sort; use the count() function)
```{r}
not_cancelled %>%
  count(dest) %>%
  arrange(desc(n))
```

Find the planes that flew the most miles. (Use count() and the wt option):
```{r}
not_cancelled %>% 
  count(tailnum, wt = distance) %>%
  arrange(desc(n))
```

How many flights left before 5am? (Using 'not_cancelled' flights subset group by day and summarise)
```{r}
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(n_early = sum(dep_time < 500),
            .groups = "drop_last")
```

What proportion of flights are delayed by more than an hour? 

[Using 'not_cancelled' dataset, a 'flights' subset group by day and summarise]
```{r}
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(hour_prop = mean(arr_delay > 60),
            .groups = "drop_last")
```

```{r}
(daily <- group_by(flights, year, month, day))
```

Create and show a new dataset called 'per_day' that is the count of flights per day using the 'summarise' function on the 'daily' dataset.

[The 'daily' dataset is the 'flights' dataset grouped by year, month, day]
```{r}
(per_day   <- summarise(daily, flights = n()))
```
Create and show a new dataset called 'per_month' that is the count of flights per month using the 'summarise' function on the 'per_day' dataset. (Note that a group is "peeled" off after doing the summarise)

[The 'per_day' dataset is summarized from the 'flights' dataset having the count of flights grouped by year, month, day]
```{r}
(per_month <- summarise(per_day, flights = sum(flights)))
```
Create and show a new dataset called 'per_year' that is the count of flights per year using the 'summarise' function on the 'per_month' dataset. (Note that a group is "peeled" off after doing the summarise)

[The 'per_month' dataset is summarized from the 'flights' dataset having the count of flights grouped by year, month]

```{r}
(per_year  <- summarise(per_month, flights = sum(flights)))
```
Show the count of all flights by changing the 'daily' dataset with an 'ungroup' function.

[Using the 'daily' dataset which is created from the flights dataset with grouping by year, month, day.]
```{r}
daily %>% 
  ungroup() %>%             # no longer grouped by date
  summarise(flights = n())  # all flights
```
Come up with another approach that will give you the number of flights by destination, i.e. not_cancelled %>% count(dest), but without using count()

[Using 'not_cancelled' dataset, a 'flights' subset]
```{r}
not_cancelled %>% 
  group_by(dest) %>%
  summarise(n = n())
```
Come up with another approach that will give you the number of flights by destination, i.e. not_cancelled %>% count(dest), and use the tally() function

[Using 'not_cancelled' dataset, a 'flights' subset]

```{r}
not_cancelled %>%
  group_by(dest) %>%
  tally()
```


Come up with another approach that will give you the miles traveled by tailnum, i.e. not_cancelled %>% count(tailnum, wt = distance), but without using count() 
 
[Using 'not_cancelled' dataset, a 'flights' subset]
```{r}
not_cancelled %>% 
  group_by(tailnum) %>%
  summarise(dist = sum(distance)) 
```

Come up with another approach that will give you the miles traveled by tailnum, i.e. not_cancelled %>% count(tailnum, wt = distance), and use the tally function
 
[Using 'not_cancelled' dataset, a 'flights' subset]
```{r}
not_cancelled %>%
  group_by(tailnum) %>%
  tally(distance)
```
Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of cancelled flights related to the average delay?
```{r}
cancel_per_day <- flights %>%
  mutate(cancelled = (is.na(arr_delay) | is.na(dep_delay))) %>%
  group_by(month, day) %>%
  summarise(
    NumCancelled = sum(cancelled),
    NumFlights = n(),
    AveDelay = mean(arr_delay, na.rm = TRUE),
    CancelProp = mean(cancelled)
  )
```

```{r}
ggplot(data = cancel_per_day) +
  geom_point(mapping = aes(x = NumFlights, y = NumCancelled))
```


```{r}
ggplot(data = cancel_per_day) +
  geom_point(mapping = aes(x = AveDelay, y = CancelProp))
```

> 5A. Which carrier has the worst delays? [Use 'not_cancelled' dataset (a 'flights' subset). ]

```{r}
not_cancelled %>%
  mutate(Delay = (arr_delay >= 0 | dep_delay >= 0)) %>%
  group_by(carrier) %>%
  summarise(
    DelayProportion = mean(Delay),
    n = n()
  ) %>%
  arrange(desc(DelayProportion))
```

> 5B. Can you disentangle the effects of bad airports vs. bad carriers? [Use 'flights' dataset. Create a Delay column based on both 'arr_delay' and 'dep_delay'. Sort destinations based on delay proportion.]

```{r}
not_cancelled %>%
  mutate(Delay = (arr_delay >= 0 | dep_delay >= 0)) %>%
  group_by(dest) %>%
  summarise(
    DelayProportion = mean(Delay),
    n = n()) %>%
  arrange(desc(DelayProportion))
```


